On the effects of partying on next-day sports performance

### 2017-07-11 writing/draft.tex
Updated mean points allowed regression table to include possible confounders for discrete indicator model. Update county --> MSA.

% \paragraph{What do we find}
We perform this analysis for both the National Basketball Association
(NBA) and Major League Baseball (MLB). 
We find that visiting a city with active nightlife the day prior to a game does have a 
statistically significant negative effect on a team's likelihood of meeting bookmakers' expectations
for both NBA and MLB. Within the NBA, we analyze specific team performance metrics such as number of points allowed and number of rebounds obtained in a game, and find that both are negatively effected by previous day visitation to a city with active nightlife. Having established a causal model, we realize a profitable betting scheme for MLB. We also perform a robustness check and see that if players rest more than
 24 hours after visiting a ``party city'', the hangover effect dissipates.

@@

We plot a heatmap which displays the logged count of co-occurences between last game
location and current game location, split by home-and-away games.
Our takeaway from this figure is that indeed there is some correlation between last game location
and next day opponent (more so when both are away games,  
and in particular this is true for west coast teams). We hypothesize that
this is due to practical limitations on flight schedules. E.g. if a team is unwilling to take a red-eye
flight, then there is little hope to play an evening west coast game and make it to the east coast for a game the very next day;
the same is \emph{not true} when flying from east to west, in which case a morning flight is reasonable.

@@

\paragraph{A continuous measure}
We also found that athletes tend to party with other A-list celebrities,
particularly musicians.\citep{jbutlerdwade} After winning the NBA finals
in 2016, the Cleveland Cavaliers were found celebrating in nightclubs.\citep{xsclub} 
This motivated a continuous measure of nightlife, based on a simple count of the number of musicians in the metropolitan statistical area (MSA) 
in which the stadium resides: if the last game was played at most 24-hours ago, use the log of the total number
of musical groups, sound recording studios, and musical publishers in county of the 
last game location, else emit zero.  We also re-scale this feature to lie within the
interval $[0,1]$ such that the magnitude of the coefficient is comparable with our
discrete indicator of nightlife.
This captures how much nightlife there is in a city since musical performances are correlated
with parties, if not an essential feature.

@@

We interpret the point-spread according to a standard guide.\citep{hrisports} 
The reason we utilize bookmaker spreads is because this measure already incorporates
up-to-date information on how well the team as been performing in said season and also
their expected (dis)-advantage relative to a particular opponent. Without a feature
describing bookmaker expectations, we would need to construct a separate model (from scratch)
which estimates the probability that one team will be victorious over another, for all
$\binom{n}{2}$ possible match-ups.

@@

\paragraph{MLB} We also obtain data for seasons 2010 through 2017; each season runs from 
April through October. There are also exactly 30 teams in the MLB, each plays 162 games per season.
We don't have all $162 \times 30 / 2 = 2,430$ games played each season, but instead only have
on average almost 2,100 games per season. We posit that the missing observations
are missing in a way that is exogenous to our model; a hand-inspection of several
reveals that the website URL was defunct.
We have a total of 12,709 games in our data-set. We again add features describing rest time and travel distance.

@@

\subsection{BLS data} We also obtained data from
\href{https://www.bls.gov/data/}{Bureau of Labor Statistics} which records the number of establishments by business type at the MSA-quarter level. We obtain this data for each metropolitan statistical area corresponding
to an NBA or MLB team from years 2010-2016. 
As a proxy for how much night-life there is in a city, we look toward the 
number of sound recording studios, musical groups, and music publishers there 
are for a particular county-year: we simply calculate the total number of establishments across all three music categories listed above. We caveat that since the Toronto Raptors and Toronto Blue Jays are located in Canada, 
we don't have BLS data for these teams.\footnote{We don't expect Toronto, Canada to be a party city.}

@@

\paragraph{Model} We use a logistic regression to estimate the likelihood of meeting
the spread as a function of whether the team partied as well as controlling for other confounders
listed above. Our basic model is as follows accounts for the rest-time since
last game, jet-lag, home-team effect, player fatigue incurred during the last game,
and the party effect.

@@

\subsubsection{Betting performance}
We also validate our MLB model by seeing how well we can perform against the
market.
\begin{figure}[!h]
  \centering
  \label{mlb bets by season}
  \includegraphics[scale=0.82]{mlb_bets_by_season}
  \caption{Betting performance by season. In each season, the 
    circled point represents the worst out of pocket expense incurred.
    We first train a model on historical data then use said model to predict on ``current year''
    data, plotting profits as a function of time over the ``current'' season. For each relevant
    game, we place a bet of \$100 if the expected value of the bet is positive after accounting for the bookmaker's cut. E.g. in the     early 2016 season,
    at its worst we are out of pocket \$750 dollars. By season end, we have     recouped all of
    our initial investment plus an additional \$11.5k. We are profitable in each season, and mention that 2015 is the ``worst'', but we still come out ahead.}
  \label{bettingperf}
\end{figure}

@@

% mpa1, tpa, tpo
\hspace{-125pt}
\begin{tabular}{@{\extracolsep{5pt}}lccc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]   & \multicolumn{3}{c}{\textit{Dependent variable:}} \\  \cline{2-4}  \\[-1.8ex] & Team Points Admitted & Team Points Admitted & Team Points Scored \\  \\[-1.8ex] & (1) & (2) & (3)\\  \hline \\[-1.8ex]   Discrete party indicator & 2.817$^{***}$ &  &  \\    & (0.712) &  &  \\    & & & \\   Continuous party measure &  & 1.942$^{***}$ & 0.377 \\    &  & (0.248) & (0.246) \\    & & & \\   Lag change in possessions &  & 0.022$^{***}$ & 0.035$^{***}$ \\    &  & (0.007) & (0.007) \\    & & & \\   Logged travel distance &  & $-$0.411$^{***}$ & 0.035 \\    &  & (0.074) & (0.073) \\    & & & \\   Number hours since last game &  & 0.012$^{***}$ & 0.020$^{***}$ \\    &  & (0.003) & (0.003) \\    & & & \\   Constant & $-$0.046 & 102.824$^{***}$ & 99.540$^{***}$ \\    & (0.091) & (0.483) & (0.478) \\    & & & \\  \hline \\[-1.8ex]  Observations & 15,139 & 44,689 & 44,689 \\  R$^{2}$ & 0.001 & 0.002 & 0.002 \\  Adjusted R$^{2}$ & 0.001 & 0.002 & 0.002 \\  Residual Std. Error & 11.137 (df = 15137) & 12.117 (df = 44684) & 12.015 (df = 44684) \\  F Statistic & 15.670$^{***}$ (df = 1; 15137) & 27.968$^{***}$ (df = 4; 44684) & 24.325$^{***}$ (df = 4; 44684) \\  \hline  \hline \\[-1.8ex]  \textit{Note:}  & \multicolumn{3}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\  \end{tabular}

\section{Conclusions and Avenues for Future Research}
\subsection{Extending the research}
\paragraph{Within the NBA} 
We're interested to explore what other factors may affect a team's propensity to suffer hangover effects beyond simply visiting a party city; e.g. are teams more likely to go out after
a win or a loss? Do teams who aren't going to make the playoffs party more? Are there demographic
factors which influence partying, e.g. are younger, unmarried players more susceptible to treatment effect? Within a game, it would be interesting to look at 
number of shots taken, and number of shots taken by 
stars as a function of previous day visitation to party cities. We could also investigate 
effects on injury risk and flagrant foul rates.

>>>>

% \paragraph{What do we find}
We perform this analysis for both the National Basketball Association
(NBA) and Major League Baseball (MLB). 
We find that visiting a city with active nightlife the day prior to a game does have a 
statistically significant negative effect on a team's likelihood of meeting bookmakers' expectations
for both NBA and MLB. Within the NBA, we analyze specific team performance metrics such as number of points allowed, and find that it is 
negatively effected by previous day visitation to a city with active nightlife. Having established a causal model, we realize a profitable betting scheme for MLB. We also perform a robustness check and see that if players rest more than
 24 hours after visiting a ``party city'', the hangover effect dissipates.

@@

The figure displays the a count of co-occurences between last game location and current game location, faceted by home and away games.
Our takeaway from this figure is that indeed there is some correlation between last game location
and next day opponent (more so when both are away games,  
and in particular this is true for west coast teams). We hypothesize that
this is due to practical limitations on flight schedules. E.g. if a team is unwilling to take a red-eye
flight, then there is little hope to play an evening west coast game and make it to the east coast for a game the very next day;
the same is \emph{not true} when flying from east to west, in which case a morning flight is reasonable.

@@

\paragraph{A continuous measure}
We also found that athletes tend to party with other A-list celebrities,
particularly musicians.\citep{jbutlerdwade} After winning the NBA finals
in 2016, the Cleveland Cavaliers were found celebrating in nightclubs.\citep{xsclub} 
This motivated a continuous measure of nightlife, based on a simple count of the number of musicians in the metropolitan statistical area (MSA) 
in which the stadium resides: if the last game was played at most 24-hours ago, use the log of the total number
of musical groups, sound recording studios, and musical publishers in the MSA of the 
last game location, else emit zero.  We also re-scale this feature to lie within the
interval $[0,1]$ such that the magnitude of the coefficient is comparable with our
discrete indicator of nightlife.
This captures how much nightlife there is in a city since musical performances are correlated
with parties, if not an essential feature.

@@

We interpret the point-spread according to a standard guide.\citep{hrisports} 
The reason we utilize bookmaker spreads is because this measure already incorporates
up-to-date information on how well the team has been performing in said season and also
their expected (dis)-advantage relative to a particular opponent. Without a feature
describing bookmaker expectations, we would need to construct a separate model (from scratch)
which estimates the probability that one team will be victorious over another, for all
$\binom{n}{2}$ possible match-ups.

@@

\paragraph{MLB} We also obtain data for seasons 2011 through 2017; each season runs from 
April through October. There are also exactly 30 teams in the MLB, each plays 162 games per season.
We don't have all $162 \times 30 / 2 = 2,430$ games played each season, but instead only have
on average almost 2,100 games per season. We posit that the missing observations
are missing in a way that is exogenous to our model; a hand-inspection of several
reveals that the website URL was defunct.
We have a total of 12,709 games in our data-set. We again add features describing rest time and travel distance.

@@

\subsection{BLS data} We also obtained data from
\href{https://www.bls.gov/data/}{Bureau of Labor Statistics} which records the number of establishments by business type at the MSA-quarter level. We obtain this data for each metropolitan statistical area corresponding
to an NBA or MLB team from years 2010-2016. 
As a proxy for how much night-life there is in a city, we look toward the 
number of sound recording studios, musical groups, and music publishers there 
are for a particular MSA and quarter: we simply calculate the total number of establishments across all three music categories listed above. We caveat that since the Toronto Raptors and Toronto Blue Jays are located in Canada, 
we don't have BLS data for these teams.\footnote{We don't expect Toronto, Canada to be a party city.}

@@

\paragraph{Model} We use a logistic regression to estimate the likelihood of meeting
the spread as a function of whether the team partied as well as controlling for other confounders
listed above. Our basic model accounts for the rest-time since
last game, jet-lag, home-team effect, player fatigue incurred during the last game, and the party effect.

@@

\subsubsection{Betting performance}
We also validate our MLB model by seeing how well we can perform against the
market.
\begin{figure}[!h]
  \centering
  \label{mlb bets by season}
  \includegraphics[scale=0.82]{mlb_bets_by_season}
  \caption{Betting performance by season. In each season, the 
    circled point represents the worst out of pocket expense incurred.
    We first train a model on historical data then use said model to predict on ``current year''
    data, plotting profits as a function of time over the ``current'' season. For each relevant
    game, we place a bet of \$100 if the expected value of the bet is positive after accounting for the bookmaker's cut. E.g. in the early 2016 season,
    at its worst we are out of pocket \$89 dollars. By season end, we have     recouped all of
    our initial investment plus an additional \$11.5k. We are profitable in each season, and mention that 2015 is the ``worst'', but we still come out ahead.}
  \label{bettingperf}
\end{figure}

@@

% mpa1, tpa, tpo
\hspace{-125pt}
\begin{tabular}{@{\extracolsep{5pt}}lccc}  \\[-1.8ex]\hline  \hline \\[-1.8ex]   & \multicolumn{3}{c}{\textit{Dependent variable:}} \\  \cline{2-4}  \\[-1.8ex] & Team Points Admitted & Team Points Admitted & Team Points Scored \\  \\[-1.8ex] & (1) & (2) & (3)\\  \hline \\[-1.8ex]   Discrete party indicator & 2.649$^{***}$ &  &  \\    & (0.718) &  &  \\    & & & \\   Continuous party measure &  & 1.833$^{***}$ & 0.447$^{*}$ \\    &  & (0.246) & (0.245) \\    & & & \\   Lag change in possessions & 1.058$^{*}$ & 0.022$^{***}$ & 0.035$^{***}$ \\    & (0.615) & (0.007) & (0.007) \\    & & & \\   Logged travel distance & $-$0.045 & $-$0.028 & $-$0.213$^{***}$ \\    & (0.038) & (0.074) & (0.074) \\    & & & \\   Number hours since last game & 0.009$^{**}$ & 0.015$^{***}$ & 0.018$^{***}$ \\    & (0.004) & (0.003) & (0.003) \\    & & & \\   Home team effect & $-$2.997$^{***}$ & $-$3.412$^{***}$ & 2.211$^{***}$ \\    & (0.218) & (0.116) & (0.116) \\    & & & \\   Constant & $-$3.429 & 101.941$^{***}$ & 100.112$^{***}$ \\    & (2.926) & (0.479) & (0.477) \\    & & & \\  \hline \\[-1.8ex]  Observations & 14,843 & 44,689 & 44,689 \\  R$^{2}$ & 0.018 & 0.021 & 0.010 \\  Adjusted R$^{2}$ & 0.017 & 0.021 & 0.010 \\  Residual Std. Error & 11.070 (df = 14831) & 12.002 (df = 44683) & 11.967 (df = 44683) \\  F Statistic & 24.021$^{***}$ (df = 11; 14831) & 196.086$^{***}$ (df = 5; 44683) & 92.809$^{***}$ (df = 5; 44683) \\  \hline  \hline \\[-1.8ex]  \textit{Note:}  & \multicolumn{3}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\  \end{tabular}    
\section{Conclusions and Avenues for Future Research}
\subsection{Extending the research}
\paragraph{Within the NBA} 
We're interested to explore what other factors may affect a team's propensity to suffer hangover effects beyond simply visiting a party city; e.g. are teams more likely to go out after
a win or a loss? Do teams who aren't going to make the playoffs party more? Are there demographic
factors which influence partying, e.g. are younger, unmarried players more susceptible to treatment effect? Within a game, it would be interesting to look at 
number of shots taken, and number of shots taken by 
stars as a function of previous day visitation to party cities. We could also investigate 
effects on injury risk and flagrant foul rates.
